Maximum Entropy Spectral Analysis: a case study

### 2021-05-07 paper/mesa.tex
Fixed typos

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ABSTRACT
\begin{abstract}
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package \texttt{memspectrum}. We characterize our implementation by performing a power spectral density analysis on syntethic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using syntethic data generated from the released spectrum by the LIGO-Virgo collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance and bias. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime.

@@

An ubiquitous method for such computation is due to Welch \cite{Welch1967} and it based on eqs.(\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
The PSD is obtained by slicing the observed realisation $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many window-corrected batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent~\cite{Lomb,Scargle} to taking the Fourier Transform of the windowed sample autocorrelation $\rho_W$, written as
\begin{equation}
    \rho_{W} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $\rho$ is the empirical autocorrelation and $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different way, each choice presenting advantages and disadvantages for the final estimate of the PSD.

@@

The maximisation of entropy, supplemented by evidence in the form of constraints to which the sought-for probability distribution must obey,
gives rise to several of the most common probability distributions commonly employed in statistics. For instance, whenever the only constraint available is the normalisation of the probability distribution (i.e. no evidence is available), the entropy is maximised by the uniform distribution. 
If we have evidence to constraint the expected value, the information entropy is maximised by the exponential distribution

@@

For this work, we are particularly interested in the multi-dimensional case. Suppose we have a vector of measurements $(x(t_1),\ldots,x(t_n)) = (x_1, \ldots, x_n)$ that we conveniently express as a single realisation of an unknown stochastic process $x(t)$ and we have information about the expectation value of the process $\mu(t)$ and on the matrix of autocovariances $C_{ij} \equiv C(t_i,t_j)$, then the MAXENT distribution is the $n$-dimensional multivariate Gaussian distribution~\cite{gregory_2005}: 
\begin{align}
    p&\left((x_1, \ldots, x_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)\,. 
\end{align}

@@

}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalised by the discrete Fourier transform base~\cite{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}\label{eq:Whittle}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{n / 2}}\exp\left(-\frac{1}{2}\sum_{ij}\tilde{x}_i S^{-1}_{ij} \tilde{x}_j \right)\,,
\end{align}
where the matrix $S_{ij} = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognise the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method\cite{prob_information_theory}
%P. M. Woodward, D. W. Fry, and W. Higinbotham, Probability and information theory, with applications to radar; 2nd ed. (Pergamon Press, Oxford, 1964), URL http://cds.cern.ch/record/2031792.
and of gravitational waves data analysis, e.g. \cite{Finn_1992, FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the autocovariance function.

@@

Before doing so, there is a technical problem to solve: the definition of entropy depends on a probability distribution, not on the PSD.
It can be shown, e.g.~\cite{AblesMESA, Bartlett}, that the variational problem can be formulated in terms of the power spectral density $S(f)$ alone by considering our signal as the result of the filtering a white noise process using a filter with transfer function $T(f)$ equal to $S(f)$\footnote{
A filter with transfer function $T(f)$ takes in input a time series $x_t$ and outputs a times series $y_t$ such that:
$$T(f) = \frac{\tilde{y}(f)}{\tilde{x}(f)}$$
where $\tilde{x}(f)$ denotes the Fourier transform of $x_t$ (and similarly for $y_t$)
}.
The difference in entropy between the input and the output time series (i.e. the entropy gain) obtained by such filter applied on white noise is:
\begin{equation}\label{eq:EntropyGain}
    \Delta H = \int_{- Ny}^{Ny}\log S(f) df\,.
\end{equation}
where $\Delta t$ is sampling rate and $Ny \equiv \frac{1}{2 \Delta t}$  is the Nyquist frequency.
Thus maximising Eq.~\eqref{eq:EntropyGain} is equivalent to maximise Eq.~\eqref{eq:entropy_continuos}.

Before maximising the entropy gain, we need to include the evidence available as a form of mathematical constraints for the assignment of $S(f)$.
This is equivalent in imposing that the variational solution $S(f)$ for the PSD matches the empirical autocorrelation.
Let us define a realisation of a stochastic process $(x_1,\ldots,x_N)$ with sample autocorrelations $\bar r_k,\,k=0,\ldots, N/2$, then the PSD must satisfy the following equation:
\begin{equation}\label{eq:MaxConstraint}
\int_{-Ny}^{Ny} S(f) e^{\imath 2 \pi f k \Delta t} df = \bar r_{k}\,.
\end{equation}

Thus, by maximizing Eq.~\eqref{eq:EntropyGain} with constraints in eq.~\eqref{eq:MaxConstraint}, we can give an estimate of the spectrum given an empirical time series.
This approach on PSD computation provides a result consistent with the epirical autocorrelation function whenever this is available and, at the same time, it does not make any assumption for the unavailable estimates for the autocorelation at large time lags.

@@

The reminder of this section is devoted to an extensive study on how to make such choice.
In Section~\ref{sec:optimizers}, we are going to define three different \textit{loss functions} to measure how well the 
algorithm is able to reproduce a known PSD.
The basic idea is to validate, as the autoregressive order considered increase, the performance of the algorithm results 
by measuring the loss function and pick, among the orders the one that yields better results.
Second, the performance of different loss functions will be assessed by performing the spectral analysis on time series with an analytical gaussian PSD, sec.~\ref{sec:arp_validation}.
In a last subsection~\ref{sec:LIGO_validation}, the same analysis is performed on syntetic data generated from the analytical spectrum released by the LIGO-Virgo collaboration.

@@

\begin{itemize}
\item \textbf{Final prediction Error} 
The first criterion is due to Akaike~\cite{Akaike1970StatisticalPI}. It was proposed that $m$ should be chosen as the 
length that minimises the error when the filter is used as a predictor, the \emph{final prediction} error (FPE): 
\begin{equation}
    FPE(m) = \mathbb{E}\left[ \left((x_t - \hat x_t) ^ 2\right) \right]
\end{equation}
with $\hat{x}_t = \sum_{i = 1}^M a_i x_{t - i}$.
Minimizing FPE is equivalent to minimizing the quantity: 
\begin{equation}
    \mathcal{L}_{\rm FPE}(m) = P_{m} \frac{N + m + 1}{N - m - 1}
\end{equation}
with $P_m$ being the estimated noise variance at order $m$, see Eq.~\eqref{eq:errorFilter1}. In the $N \to \infty$ limit, 
remembering $m_{max} \sim 2N / \log(2N)$, Akaike's loss function is equivalent to the minimization of the variance $P_m$ of the white noise of the underlying $AR(p)$ model.

@@

We then apply Burg's method choosing in turn each of the loss functions on the ensemble of simulated time series, 
thus obtaining an ensemble of PSD estimates. Through these ensembles, we characterise statistically the performance of each loss function.
The disagreement between the PSD $S_i(f)$ estimated from the $i$-th simulated time series and the target PSD $S(f)$ is measured via 
the \textit{frequency-averaged relative error} $r_i$:
\begin{equation}\label{eq:freq_error}
	r_i = \frac{1}{N_f}\sum_{f_j=\{0, \hdots , Ny\}} \frac{|S_i(f_j) - S(f_j)|}{S(f_j)}
\end{equation}
where $Ny$ is the Nyquist frequency of the time series and $N_f$ is the number of the discrete frequencies the PSD is evaluated at.

@@

By looking at Fig.~\ref{fig:optcomparison} (red series), we note that the AR orders obtained with 
FPE are clustered in a very small region. This is also a desirable property: FPE, in fact, provides a stable 
estimation of the ar order, which does not affect much the reconstruction error.
We conclude that the FPE shows good quality reconstruction for the spectrum and very desirable stability properties, its estimate for filter's length $m$ is clustered in a region where there is no dependence of error on $m$. 
\paragraph{Optimum Bayes Decision Rule (OBD)}
The second loss function we cosider is OBD and the results are summerised in Fig.~\ref{fig:OBDmean}. As for the FPE case, they show a good agreement between the average over the reconstructions and the true spectrum.
Qualitatively the same behaviour of FPE is observed: a good quality reconstruction at the intermediate and high frequencies with a narrow $90\%$ confidence level as well as a degrading performance at low frequencies.
However, when looking at the error, the disagreement of this method is found to be larger compared to FPE: 
in the worst case, the error can be as large as a factor of $2$ compared to FPE.

@@

When averaging the different PSD estimates, the noise in each spectrum cancels, as expected from the Gaussian nature of the AR process. 
This implies, in a sense, that each estimate of the spectrum is independent of any other, as suggested by the huge variance in the residuals. This lack of stability is not a good property for the estimation of the PSD from a single realisation of the time-series, however, thanks to the averaging out of errors, this estimator seems optimal in the case of repeatable experiments and ensemble-averages.

@@

However, we also note that the when inferring PSDs from a single time-series realisation, FPE provide the lowest averaged error over all frequencies, while CAT can reach errors $5$ times larger (see again Fig.~\ref{fig:optcomparison}).
Hence, while CAT is the loss function that minimizes errors in the low-frequency end of the spectrum, FPE obtains the best overall accuracy.

The conclusion is clear: even if in some cases CAT is more accurate when taking the average over several realisations of the underlying process, 
FPE guarantees that the single estimation is more faithful.
As in any common situation we cannot perform such averaging over different realization of the same time series, we must prefer FPE over CAT (let alone OBD, which even though qualitatively similar to FPE has worse performance).
However, if we indeed can measure the PSD by averaging over different time series, using CAT as a loss function is the best choice. In this sense we retain CAT to provide the best, and most similar in spirit, alternative to the commonly employed Welch estimation method whenever ensemble-averages are needed and justified.

\subsection{Choice of the loss function: LIGO Spectrum} \label{sec:LIGO_validation}
We continue our characterisation of the various loss functions considered in this work, by investigating the 
reconstruction of a specific, known, power spectrum: that is the Advanced LIGO design sensitivity theoretical spectral curve\cite{LIGO_PSDs}.
For this analysis, we generate $N = 500$ time series of $40960$ points each, sampled with a sampling rate of $\SI{2048}{Hz}$, hence we fix the duration
of the time-series to $\SI{20}{s}$. The chosen length is convenient to capture fairly accurately the low-frequency features of the LIGO PSD.
We report our findings in Figs.~\ref{fig:ligospectrum}~and~\ref{fig:LigoOrderError}.

@@

Fig.~\ref{fig:LigoOrderError} shows the distribution of recovered AR orders $m$ against the relative frequency-averaged error. The behaviour
of the three loss functions is very similar to what found in the Gaussian PSD case (compare it with fig.~\ref{fig:optcomparison}): ODB infers the smallest orders and gives average errors around $20$\%, FPE consistently estimates orders of a few hundreds and shows the smallest errors $\sim 15$ \% while CAT does not show any preference towards any AR order and displays wildly varying errors. Yet again, when the PSD is averaged over multiple realization of the time, CAT is able to capture the spectrum very precisely. In fact, even in presence of very sharp spectral features, CAT reconstruction seems to be almost perfectly coincident with each of them.
Hence, also the study of simulated LIGO data seems to indicate that whenever and wherever ensemble-averaged PSD are necessary, CAT is the optimal choice of loss function. However, on a single time-series realisation, FPE is the more robust choice.

Let us summarise some key general conclusions:
\begin{itemize}
	\item there are no reasons to prefer OBD over CAT or FPE;
	\item if we have one single realization for the process, we recommend the use of FPE, that would get the best resolution possible. In this situation, CAT would provide spurious and unreliable results, with large error;
	\item in the case of several realizations of the same process, CAT ensemble-average properties provide very a precise spectral estimation.\end{itemize}

@@

It can be seen that OBD and FPE loss functions show very similar behaviour. They are {\it very} realiable in capturing the correct AR order up to a certain threshold ($p\sim 100$ for OBD and $p\sim 200$ for FPE). For any AR process of order higher than the threshold, the optimization underestimates badly the actual value of $p$.
For this reason, FPE and OBD seem reliable only for relatively simple AR processes. For more complicated processes, they seem to ``underfit'' the problem (i.e. they output a model simpler that those required).

@@

We simulate data\footnote{This is to ensure that we have a baseline PSD to compare the data with} from the PSD around the event GW150914 and we employ both Welch's method and MESA to estimate the spectrum.
We vary the length of the data used for the estimation: this is also useful to assess how the computation depends on the data available. We set the total observation time $T = 1, 5, 10, 100, 1000 \SI{}{s}$
For the MESA algorithm, we choose the FPE loss function. For the Welch algorithm, we employ a Tukey window with a parameter $\alpha$ 
equal to 0.4, an overlap fraction of $1/2$ for the segments and a length of 
segments $L = 512, 1024, 2048, 8192, 32768$ points, depending on the observation time.
In all cases, the sampling rate is set to $\SI{4096}{Hz}$.
For the Welch algorithm, we use the standard implementation provided by the python library \texttt{scipy}~\cite{numpy,scipy}.
The results from both methods are summarised in  Figs.~\ref{fig:MESA_LIGO_data} and~\ref{fig:welch_LIGO_data} respectively.

@@

\begin{figure}[!t]
	\caption{Comparison between analytic (dashed line) and estimated (red line) spectrum. The estimation is perfomed with Maximum Entropy method on \textit{synthetic} data. The estimation is performed with an increasing observation time $T = 1, 5, 10, 100, \SI{1000}{s}$.}
	\label{fig:MESA_LIGO_data}
	\includegraphics{Images/comparison_LVC_data/comparison_LVC_data_MESA.pdf}
\end{figure}

\begin{figure}[!t]
	\caption{Comparison between analytic (dashed line) and estimated (green line) spectrum. The estimation is perfomed with Welch's method on \textit{synthetic} data. The estimation is performed with an increasing observation time $T = 1, 5, 10, 100, \SI{1000}{s}$.}
	\label{fig:welch_LIGO_data}
	\includegraphics{Images/comparison_LVC_data/comparison_LVC_data_Welch.pdf}
\end{figure}

@@

We now assess the accuracy of the forecasting of new observations of the time series. 
Based on actual data, we try to predict future values as described in sec.~\ref{sec:forecasting} (of course the MESA is performed with CAT loss function). We produce $N =100$ indepedent predictions and we compute the median as well as the $90\%$ confidence interval. This is compared with the actual measured temparature values. The predictions spans a two years range of time. We report our results in Fig.~\ref{fig:temp_spectrum}.

We note the observed difference is always well included in the $90\%$ confidence interval: the forecasting predictions seem reliable.
On the other hand, the confidence interval is pretty large (almost as large as $\SI{15}{K}$), thus making ``easy" for the actual data to fit the predictions.
Indeed, the prediction model, while suitable for spectral estimation, is nothing more than a linear regression (plus noise term). Such a simple model hardly catches the complex trend in the variability of the temperature daily trend during a year. For more precise predictions, probably one should consider nonlinear regression model, tapping into the wealth of non linear predictors offered by the field of Deep Lerning.

@@

Despite poor predictions on long timescales, the AR process obtained with MESA still can be useful on short timescales. Indeed, a precise prediction of the strain time series can be beneficial in the detection of anomalies in the data and, eventually, their removal. The predictions can form an expected baseline for the strain; any anomaly (i.e. a glitch or even a transient of physical origin) can show up as a large departure from expected trend.
Moreover, if a glitch is detected, its shape can be estimated (as well as the confidence level) by subracting the expected signal with the actual signal.
This can work: a typical glitch~\cite{Zevin_2017} can last as long as $\SI{0.2}{s}$, close to the time scale over which the forecasting is reliable.

>>>>

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ABSTRACT
\begin{abstract}
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package \texttt{memspectrum}. We characterize our implementation by performing a power spectral density analysis on synthetic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using synthetic data generated from the released spectrum by the LIGO-Virgo collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance and bias. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime.

@@

An ubiquitous method for such computation is due to Welch \cite{Welch1967} and it based on eqs.(\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
The PSD is obtained by slicing the observed realization $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many window-corrected batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent~\cite{Lomb,Scargle} to taking the Fourier Transform of the windowed sample autocorrelation $\rho_W$, written as
\begin{equation}
    \rho_{W} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $\rho$ is the empirical autocorrelation and $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different way, each choice presenting advantages and disadvantages for the final estimate of the PSD.

@@

The maximisation of entropy, supplemented by evidence in the form of constraints to which the sought-for probability distribution must obey,
gives rise to several of the most common probability distributions commonly employed in statistics. For instance, whenever the only constraint available is the normalization of the probability distribution (i.e. no evidence is available), the entropy is maximised by the uniform distribution. 
If we have evidence to constraint the expected value, the information entropy is maximised by the exponential distribution

@@

For this work, we are particularly interested in the multi-dimensional case. Suppose we have a vector of measurements $(x(t_1),\ldots,x(t_n)) = (x_1, \ldots, x_n)$ that we conveniently express as a single realization of an unknown stochastic process $x(t)$ and we have information about the expectation value of the process $\mu(t)$ and on the matrix of autocovariances $C_{ij} \equiv C(t_i,t_j)$, then the MAXENT distribution is the $n$-dimensional multivariate Gaussian distribution~\cite{gregory_2005}: 
\begin{align}
    p&\left((x_1, \ldots, x_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)\,. 
\end{align}

@@

}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalized by the discrete Fourier transform base~\cite{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}\label{eq:Whittle}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{n / 2}}\exp\left(-\frac{1}{2}\sum_{ij}\tilde{x}_i S^{-1}_{ij} \tilde{x}_j \right)\,,
\end{align}
where the matrix $S_{ij} = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognize the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method\cite{prob_information_theory}
%P. M. Woodward, D. W. Fry, and W. Higinbotham, Probability and information theory, with applications to radar; 2nd ed. (Pergamon Press, Oxford, 1964), URL http://cds.cern.ch/record/2031792.
and of gravitational waves data analysis, e.g. \cite{Finn_1992, FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the autocovariance function.

@@

Before doing so, there is a technical problem to solve: the definition of entropy depends on a probability distribution, not on the PSD.
It can be shown, e.g.~\cite{AblesMESA, Bartlett}, that the variational problem can be formulated in terms of the power spectral density $S(f)$ alone by considering our signal as the result of the filtering a white noise process using a filter with transfer function $T(f)$ equal to $S(f)$\footnote{
A filter with transfer function $T(f)$ takes in input a time series $x_t$ and outputs a times series $y_t$ such that:
$$T(f) = \frac{\tilde{y}(f)}{\tilde{x}(f)}$$
where $\tilde{x}(f)$ denotes the Fourier transform of $x_t$ (and similarly for $y_t$)
}.
The difference in entropy between the input and the output time series (i.e. the entropy gain) obtained by such filter applied on white noise is:
\begin{equation}\label{eq:EntropyGain}
    \Delta H = \int_{- Ny}^{Ny}\log S(f) df\,.
\end{equation}
where $\Delta t$ is sampling rate and $Ny \equiv \frac{1}{2 \Delta t}$  is the Nyquist frequency.
Thus maximising Eq.~\eqref{eq:EntropyGain} is equivalent to maximize Eq.~\eqref{eq:entropy_continuos}.

Before maximizing the entropy gain, we need to include the evidence available as a form of mathematical constraints for the assignment of $S(f)$.
This is equivalent in imposing that the variational solution $S(f)$ for the PSD matches the empirical autocorrelation.
Let us define a realization of a stochastic process $(x_1,\ldots,x_N)$ with sample autocorrelations $\bar r_k,\,k=0,\ldots, N/2$, then the PSD must satisfy the following equation:
\begin{equation}\label{eq:MaxConstraint}
\int_{-Ny}^{Ny} S(f) e^{\imath 2 \pi f k \Delta t} df = \bar r_{k}\,.
\end{equation}

Thus, by maximizing Eq.~\eqref{eq:EntropyGain} with constraints in eq.~\eqref{eq:MaxConstraint}, we can give an estimate of the spectrum given an empirical time series.
This approach on PSD computation provides a result consistent with the empirical autocorrelation function whenever this is available and, at the same time, it does not make any assumption for the unavailable estimates for the autocorelation at large time lags.

@@

The reminder of this section is devoted to an extensive study on how to make such choice.
In Section~\ref{sec:optimizers}, we are going to define three different \textit{loss functions} to measure how well the 
algorithm is able to reproduce a known PSD.
The basic idea is to validate, as the autoregressive order considered increase, the performance of the algorithm results 
by measuring the loss function and pick, among the orders the one that yields better results.
Second, the performance of different loss functions will be assessed by performing the spectral analysis on time series with an analytical Gaussian PSD, sec.~\ref{sec:arp_validation}.
In a last subsection~\ref{sec:LIGO_validation}, the same analysis is performed on synthetic data generated from the analytical spectrum released by the LIGO-Virgo collaboration.

@@

\begin{itemize}
\item \textbf{Final prediction Error} 
The first criterion is due to Akaike~\cite{Akaike1970StatisticalPI}. It was proposed that $m$ should be chosen as the 
length that minimizes the error when the filter is used as a predictor, the \emph{final prediction} error (FPE): 
\begin{equation}
    FPE(m) = \mathbb{E}\left[ \left((x_t - \hat x_t) ^ 2\right) \right]
\end{equation}
with $\hat{x}_t = \sum_{i = 1}^M a_i x_{t - i}$.
Minimizing FPE is equivalent to minimizing the quantity: 
\begin{equation}
    \mathcal{L}_{\rm FPE}(m) = P_{m} \frac{N + m + 1}{N - m - 1}
\end{equation}
with $P_m$ being the estimated noise variance at order $m$, see Eq.~\eqref{eq:errorFilter1}. In the $N \to \infty$ limit, 
remembering $m_{max} \sim 2N / \log(2N)$, Akaike's loss function is equivalent to the minimization of the variance $P_m$ of the white noise of the underlying $AR(p)$ model.

@@

We then apply Burg's method choosing in turn each of the loss functions on the ensemble of simulated time series, 
thus obtaining an ensemble of PSD estimates. Through these ensembles, we characterize statistically the performance of each loss function.
The disagreement between the PSD $S_i(f)$ estimated from the $i$-th simulated time series and the target PSD $S(f)$ is measured via 
the \textit{frequency-averaged relative error} $r_i$:
\begin{equation}\label{eq:freq_error}
	r_i = \frac{1}{N_f}\sum_{f_j=\{0, \hdots , Ny\}} \frac{|S_i(f_j) - S(f_j)|}{S(f_j)}
\end{equation}
where $Ny$ is the Nyquist frequency of the time series and $N_f$ is the number of the discrete frequencies the PSD is evaluated at.

@@

By looking at Fig.~\ref{fig:optcomparison} (red series), we note that the AR orders obtained with 
FPE are clustered in a very small region. This is also a desirable property: FPE, in fact, provides a stable 
estimation of the ar order, which does not affect much the reconstruction error.
We conclude that the FPE shows good quality reconstruction for the spectrum and very desirable stability properties, its estimate for filter's length $m$ is clustered in a region where there is no dependence of error on $m$. 
\paragraph{Optimum Bayes Decision Rule (OBD)}
The second loss function we consider is OBD and the results are summarized in Fig.~\ref{fig:OBDmean}. As for the FPE case, they show a good agreement between the average over the reconstructions and the true spectrum.
Qualitatively the same behaviour of FPE is observed: a good quality reconstruction at the intermediate and high frequencies with a narrow $90\%$ confidence level as well as a degrading performance at low frequencies.
However, when looking at the error, the disagreement of this method is found to be larger compared to FPE: 
in the worst case, the error can be as large as a factor of $2$ compared to FPE.

@@

When averaging the different PSD estimates, the noise in each spectrum cancels, as expected from the Gaussian nature of the AR process. 
This implies, in a sense, that each estimate of the spectrum is independent of any other, as suggested by the huge variance in the residuals. This lack of stability is not a good property for the estimation of the PSD from a single realization of the time-series, however, thanks to the averaging out of errors, this estimator seems optimal in the case of repeatable experiments and ensemble-averages.

@@

However, we also note that the when inferring PSDs from a single time-series realization, FPE provide the lowest averaged error over all frequencies, while CAT can reach errors $5$ times larger (see again Fig.~\ref{fig:optcomparison}).
Hence, while CAT is the loss function that minimizes errors in the low-frequency end of the spectrum, FPE obtains the best overall accuracy.

The conclusion is clear: even if in some cases CAT is more accurate when taking the average over several realizations of the underlying process, 
FPE guarantees that the single estimation is more faithful.
As in any common situation we cannot perform such averaging over different realization of the same time series, we must prefer FPE over CAT (let alone OBD, which even though qualitatively similar to FPE has worse performance).
However, if we indeed can measure the PSD by averaging over different time series, using CAT as a loss function is the best choice. In this sense we retain CAT to provide the best, and most similar in spirit, alternative to the commonly employed Welch estimation method whenever ensemble-averages are needed and justified.

\subsection{Choice of the loss function: LIGO Spectrum} \label{sec:LIGO_validation}
We continue our characterization of the various loss functions considered in this work, by investigating the 
reconstruction of a specific, known, power spectrum: that is the Advanced LIGO design sensitivity theoretical spectral curve\cite{LIGO_PSDs}.
For this analysis, we generate $N = 500$ time series of $40960$ points each, sampled with a sampling rate of $\SI{2048}{Hz}$, hence we fix the duration
of the time-series to $\SI{20}{s}$. The chosen length is convenient to capture fairly accurately the low-frequency features of the LIGO PSD.
We report our findings in Figs.~\ref{fig:ligospectrum}~and~\ref{fig:LigoOrderError}.

@@

Fig.~\ref{fig:LigoOrderError} shows the distribution of recovered AR orders $m$ against the relative frequency-averaged error. The behaviour
of the three loss functions is very similar to what found in the Gaussian PSD case (compare it with fig.~\ref{fig:optcomparison}): ODB infers the smallest orders and gives average errors around $20$\%, FPE consistently estimates orders of a few hundreds and shows the smallest errors $\sim 15$ \% while CAT does not show any preference towards any AR order and displays wildly varying errors. Yet again, when the PSD is averaged over multiple realization of the time, CAT is able to capture the spectrum very precisely. In fact, even in presence of very sharp spectral features, CAT reconstruction seems to be almost perfectly coincident with each of them.
Hence, also the study of simulated LIGO data seems to indicate that whenever and wherever ensemble-averaged PSD are necessary, CAT is the optimal choice of loss function. However, on a single time-series realization, FPE is the more robust choice.

Let us summarize some key general conclusions:
\begin{itemize}
	\item there are no reasons to prefer OBD over CAT or FPE;
	\item if we have one single realization for the process, we recommend the use of FPE, that would get the best resolution possible. In this situation, CAT would provide spurious and unreliable results, with large error;
	\item in the case of several realizations of the same process, CAT ensemble-average properties provide very a precise spectral estimation.\end{itemize}

@@

It can be seen that OBD and FPE loss functions show very similar behaviour. They are {\it very} reliable in capturing the correct AR order up to a certain threshold ($p\sim 100$ for OBD and $p\sim 200$ for FPE). For any AR process of order higher than the threshold, the optimization underestimates badly the actual value of $p$.
For this reason, FPE and OBD seem reliable only for relatively simple AR processes. For more complicated processes, they seem to ``underfit'' the problem (i.e. they output a model simpler that those required).

@@

We simulate data\footnote{This is to ensure that we have a baseline PSD to compare the data with} from the PSD around the event GW150914 and we employ both Welch's method and MESA to estimate the spectrum.
We vary the length of the data used for the estimation: this is also useful to assess how the computation depends on the data available. We set the total observation time $T = 1, 5, 10, 100, 1000 \SI{}{s}$
For the MESA algorithm, we choose the FPE loss function. For the Welch algorithm, we employ a Tukey window with a parameter $\alpha$ 
equal to 0.4, an overlap fraction of $1/2$ for the segments and a length of 
segments $L = 512, 1024, 2048, 8192, 32768$ points, depending on the observation time.
In all cases, the sampling rate is set to $\SI{4096}{Hz}$.
For the Welch algorithm, we use the standard implementation provided by the python library \texttt{scipy}~\cite{numpy,scipy}.
The results from both methods are summarized in  Figs.~\ref{fig:MESA_LIGO_data} and~\ref{fig:welch_LIGO_data} respectively.

@@

\begin{figure}[!t]
	\caption{Comparison between analytic (dashed line) and estimated (red line) spectrum. The estimation is performed with Maximum Entropy method on \textit{synthetic} data. The estimation is performed with an increasing observation time $T = 1, 5, 10, 100, \SI{1000}{s}$.}
	\label{fig:MESA_LIGO_data}
	\includegraphics{Images/comparison_LVC_data/comparison_LVC_data_MESA.pdf}
\end{figure}

\begin{figure}[!t]
	\caption{Comparison between analytic (dashed line) and estimated (green line) spectrum. The estimation is performed with Welch's method on \textit{synthetic} data. The estimation is performed with an increasing observation time $T = 1, 5, 10, 100, \SI{1000}{s}$.}
	\label{fig:welch_LIGO_data}
	\includegraphics{Images/comparison_LVC_data/comparison_LVC_data_Welch.pdf}
\end{figure}

@@

We now assess the accuracy of the forecasting of new observations of the time series. 
Based on actual data, we try to predict future values as described in sec.~\ref{sec:forecasting} (of course the MESA is performed with CAT loss function). We produce $N =100$ independent predictions and we compute the median as well as the $90\%$ confidence interval. This is compared with the actual measured temperature values. The predictions spans a two years range of time. We report our results in Fig.~\ref{fig:temp_spectrum}.

We note the observed difference is always well included in the $90\%$ confidence interval: the forecasting predictions seem reliable.
On the other hand, the confidence interval is pretty large (almost as large as $\SI{15}{K}$), thus making ``easy" for the actual data to fit the predictions.
Indeed, the prediction model, while suitable for spectral estimation, is nothing more than a linear regression (plus noise term). Such a simple model hardly catches the complex trend in the variability of the temperature daily trend during a year. For more precise predictions, probably one should consider nonlinear regression model, tapping into the wealth of non linear predictors offered by the field of Deep Learning.

@@

Despite poor predictions on long timescales, the AR process obtained with MESA still can be useful on short timescales. Indeed, a precise prediction of the strain time series can be beneficial in the detection of anomalies in the data and, eventually, their removal. The predictions can form an expected baseline for the strain; any anomaly (i.e. a glitch or even a transient of physical origin) can show up as a large departure from expected trend.
Moreover, if a glitch is detected, its shape can be estimated (as well as the confidence level) by subtracting the expected signal with the actual signal.
This can work: a typical glitch~\cite{Zevin_2017} can last as long as $\SI{0.2}{s}$, close to the time scale over which the forecasting is reliable.
